---
title: "**SC2P: Two-phase differential expression for single-cell RNA-seq**"
author: "**Zhijin Wu, Department of Biostatistics, Brown University**"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{SC2P: Two-phase differential expression for single-cell RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(knitr)
```


`SC2P` is a package designed for testing differential expression (DE) for data from 
single-cell RNA-seq experiment. It provides functionalities for testing DE in two phases:
(1) phase transition (difference in the probabilities of being expressed); and 
(2) magnitude tuning (difference in the levels of expression once the gene is on). 


## 1. Introduction

Single-cell RNA-sequencing (scRNA-seq) has emerged recently as a 
powerful technology to investigate transcriptomic variation 
at the individual cell level. Compared to traditional ***bulk*** RNA-seq, scRNA-seq reveals much detailed information for inter-cellular heterogeneities. 
scRNA-seq data show clear evidence of binary status of transcription, which we refer to as ***phases*** in transcription: **Phase I** corresponds to 
low level non-specific transcription (for example, as a result of random initiation), and **Phase II** corresponds to targeted specific transcription. The regulation of transcription includes a phase transition between Phase I to Phase II, as well as continuous regulation within Phase II. Both are important regulatory mechanisms that need to be identified in the DE analysis. 

The **SC2P** package identify DE genes in two phases seperately. It implements a rigorous statistical method to determine the phases for all genes in all cells in a data-driven way, with consideration of cell- and gene-specific characteristics. Compared with methods using an *ad hoc* thresholds to determine phases, SC2P achieves better sensitivity and accuracy. 



## 2. Use SC2P

SC2P starts from a count matrix for gene expressionsm, and a data frame for cell information. In the count matrix, each row corresponds to a gene and each column corresponds to a cell. Each row of the cell information data frame contains the annotation for a cell. The number of columns of the count matrix and the number of rows for the cell data frame must match. 
Below we will use a small part of a public human brain dataset (GSE67835) to illustrate the workflow of SC2P.  The data is distributed with SC2P as `brain_scRNAseq`.


**1. Load library and example data**

```{r, message=FALSE, warning=FALSE}
library(SC2P)
data(brain_scRNAseq)
```


**2. Create an `ExpressionSet` object out of the count matrix and cell information data frame.** 


```{r, message=TRUE, warning=FALSE}
colnames(Y) <- rownames(design)
phenoData <- new("AnnotatedDataFrame", data=design)
eset <- ExpressionSet(assayData=Y, phenoData=phenoData)
eset
```

**3.  estimate the phases for all genes in all cells using `eset2Phase` function.** This returns an object of `sc2pSet`. 

```{r, message=TRUE, warning=FALSE}
data <- eset2Phase(eset)
data
```

The phase estimation result can be visualized using `zyPlot` function, which plots the posterior probability of being expressed versus log expression. Two groups will be shown in different colors. 
```{r, message=TRUE, warning=FALSE, fig.width=4, fig.height=4, fig.align="center"}
zyPlot(rownames(data)[1], data, group.name="celltype")
```



**4. test DE in two phases using `twoPhaseDE` function.**  

Here 
- `design` is a character vector of variable names in pData(norm2) to provide regression covariates in the DE test. 
- `test.which` is an integer that points to the location of the to-be-tested binary variable in argument `design`.
- `offset` indicates the method to compute normalization factor. 


```{r, message=TRUE, warning=FALSE}
de.sc2p <- twoPhaseDE(data, design="celltype", test.which=1, offset="sf")
```

The function returns a data frame, each row is for a gene. The rows are sorted by the gene name, not significance level. 


**5. Top ranked genes can be obtained using `topGene` function, and visualized by `visGene` function.**

 There are differnt options in `topGene`: 

- To list the top 5 DE genes in phase I:
```{r, message=TRUE, warning=FALSE}
 topGene(de.sc2p, phase=1, number=5)
```

- To list the top 5 DE genes in phase II:
```{r, message=TRUE, warning=FALSE}
 topGene(de.sc2p, phase=2, number=5)
```
- To list the top 5 DE genes in both phases:

```{r, message=TRUE, warning=FALSE}
topGene(de.sc2p, phase="both", number=5)
```



To visualize expression distribution using `visGene` function:

- Top DE gene in phase I
```{r, message=TRUE, warning=FALSE, fig.width=4, fig.height=4, fig.align="center"}
visGene(topGene(de.sc2p, 1)$Gene.name[1], data, group.name="celltype")
```


-  Top DE gene in phase II
```{r, message=TRUE, warning=FALSE, fig.width=4, fig.height=4,fig.align="center"}
visGene(topGene(de.sc2p, 2)$Gene.name[1], data, group.name="celltype")
```


## 3. Session Info
```{r}
sessionInfo()
```
